Density-dependent selection and the limits of relative fitness

### 2018-04-26 manuscript.tex
l

Relative fitness is often parameterized in terms of selection coefficients which represent the advantages of different types relative to each other. For instance, in continuous time $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$. Assuming that only $i$ and $j$ are present, the change in frequency can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i (1-p_i). \label{eq:canonicalcont}
\end{equation}
Thus, if $r_i$ and $r_j$ are constant, the frequency of the $i$ type will grow logistically with a constant rate parameter $s$. We then say that selection is independent of frequency and density. The discrete time case is more complicated. Defining the selection coefficient by $W_i=(1+s)W_j$, and again assuming $i$ and $j$ are the only types present, we have  
\begin{equation}
\Delta p_i=\frac{W_i-W_j}{\overline{W}} p_i (1-p_i)=\frac{s}{1+sp_i} p_i (1-p_i). \label{eq:canonicaldisc}
\end{equation}
Hence, even in the simplest case that $W_i$ and $W_j$ are constant, selection is frequency-dependent in discrete time (note that this frequency dependence is negligible when $s$ is small compared to $1$; see \citealt{frank_2011}). We will refer to both the continuous and discrete time selection equations \eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} throughout this paper, but the simpler continuous time case will be our point of comparison for the rest of this section.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} Phase diagram for the densities of two types $n_1$ and $n_2$ undergoing selection. (a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_1>K_2$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

@@

In the special case that all types are competitively equivalent (identical $c_i$), Eq.~\eqref{eq:master} takes a simpler form,
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
\end{equation}
This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these. Total population density thus grows according to
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

\subsection*{Density-dependent selection in the variable-density lottery}

We now outline the basic properties of selection on $b$, $c$ and $d$. The birth and mortality rates $b$ and $d$ are the traits which regulate density; $b$ controls the fraction of unoccupied territories that are contested, while $d$ controls adult mortality. Competitive ability $c$ does not regulate density since it only affects the relative likelihood for each type to win a contested territory. Thus, selection between types which only differ in $c$ occurs without causing $N$ to change (Eq.~\eqref{eq:Nmaster} shows this formally).

Selection in our variable density lottery model is density-dependent. This means that when two types are present, the discrete-time selection factor $(W_i-W_j)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} may depend on $N$. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because most territories are claimed without contest at low density, whereas at high density few unoccupied territories are available to be contested.

To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
\begin{equation}
\frac{\Delta_+ n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:bdensitydependence}
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. It is then clear that $d$-selection is independent of density, because the density-dependent factor $1 + \frac{\Delta_+ n_i}{n_i}$ in Eq.~\eqref{eq:absfit} is the same for types that differ in $d$ only. On the other hand, the strength of $b$-selection declines with density because the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

@@

In this section we discuss the response of density to selection, comparing our variable-density lottery to previous studies of density-regulated populations. Our focus will be on the competitive ability trait $c$, because, as mentioned in the previous section, $c$-selection has no effect on population density even though it affects fitness under crowded conditions. Since this behavior is quite different from selection on more commonly used crowding-sensitivity parameters like $K$ in the logistic model, we start by revisiting MacArthur's $K$-selection argument \citep{macarthur_1967}.

@@

By contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}. As a result, the density-independent Eq.~\eqref{eq:canonicaldisc} holds in demographic equilibrium even though $c$-selection is density-dependent.

@@

For Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_i-W_j)/\overline{W}$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$. Eq.~\eqref{eq:canonicaldisc} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model. Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require selection to be density-dependent, and also density to be changing. This can obviously occur if density-dependent selection is occurring in a population far from demographic equilibrium. In this case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonicaldisc} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize how Eq.~\eqref{eq:canonicalcont} breaks down in a simpler continuous-time birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
\end{equation}
Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence} has the same value $d_i-1$ at the beginning and end of a $b$-sweep. During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps.

If selection acts simultaneously on more than one trait in our variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}). This can produce behavior analogous to selection on $\delta$ in Fig.~\ref{fig:strengthofselection}.

\begin{figure}
\centering
\includegraphics[scale=0.8]{multiple.pdf}
\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in our variable-density lottery model as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep because $d$ is density-regulating and $c$ is density-dependent.}
\end{figure}

@@

Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet, despite the differences between $b$, $c$ and $d$, in a constant environment selection that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly. When strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:multiple}), the constant-$s$ approximation breaks down. The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}. The combination of both is necessary to threaten the constant-$s$ approximation. Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

Selection in the variable-density lottery is quite different from classical density-dependent selection (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection''). In the latter, only one life-history stage is represented, and the effects of crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation). Selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients ($\alpha$-selection). These are both ``$\delta$-like''  in the sense that selection depends on density and also causes density to change ($\delta$ is defined in Eq.~\eqref{eq:simplebirthdeath}). Strong selection is therefore sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.

@@

Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature. Even if reproductive excesses are ubquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent. For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent. However, this conclusion is sensititive to the assumptions of well-mixed resource competition models. Spatial localization of consumable resources (e.g. for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \cite{bolker_99}. In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).

Moreover, even in well mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference. Interference competition can dramatically alter the dynamics of resource exploitation \cite{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection. For instance, sexual selection can be viewed as a form of relative interference competition between genotypes. Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like. Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important. Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \cite{gill_1974}. However, as explained above, $\alpha$-selection does transiently affect population density and can therefore be regarded as $\delta$-like.

The above findings underscore that the most serious threat to the constant-$s$ approximation arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e. as a result of a temporally-variable environment. While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the constant-$s$ approximation, they require strong selection that is both density-dependent and affects a density-regulating trait (and even then the constant-$s$ approximation may hold). In contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig. 9.5 in \cite{otto_2011}; Fig. 5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness. Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2) lottery models in general are mathematically similar to the Wright-Fisher model, which should facilitate the analysis of genetic drift when $N$ is unstable.

@@

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types. To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, with low $l_1\ll 1$ and high density $l_2\approx L\gg 1$ respectively. Since $L$ is large, uncontested territories make up a negligible fraction of the total. The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This behavior is clearly pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus,   predicted growth or decline would depend on an arbitrary assignment of rarity.

@@

Eq.~\eqref{eq:cvr} reveals two key points. First, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation. In this case, the right hand side of Eq.~\eqref{eq:cvr} is approximately equal to $C^{1/2}\left(\frac{1}{L-l_i}+1-\frac{1}{C}\right)^{1/2}$, which is small for both low and high nonfocal densities. The worst case scenario occurs when $L-l_i$ is of order one, and it can be directly verified that Eq.~\eqref{eq:meanfieldr} is then still a good approximation (see Fig.~\ref{fig:approx_details}). Second, if some of the $c_j$ are much larger than the others, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large. Specifically, in the presence of a rare,  strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the right hand side of Eq. \eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}. Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when the are large differences in $c$.

>>>>

Relative fitness is often parameterized in terms of selection coefficients which represent the advantages of different types relative to each other. For instance, in continuous time $s=r_2-r_1$ is the selection coefficient of type $2$ relative to type $1$. Assuming that only $2$ and $1$ are present, the change in frequency can be written as
\begin{equation}
\frac{d p_2}{dt}=s p_2 (1-p_2). \label{eq:canonicalcont}
\end{equation}
Thus, if $r_1$ and $r_2$ are constant, the frequency of the second type will grow logistically with a constant rate parameter $s$. We then say that selection is independent of frequency and density. The discrete time case is more complicated. Defining the selection coefficient by $W_2=(1+s)W_1$, and again assuming $1$ and $2$ are the only types present, we have  
\begin{equation}
\Delta p_2=\frac{W_2-W_1}{\overline{W}} p_2 (1-p_2)=\frac{s}{1+sp_2} p_2 (1-p_2). \label{eq:canonicaldisc}
\end{equation}
Hence, even in the simplest case that $W_1$ and $W_2$ are constant, selection is frequency-dependent in discrete time (note that this frequency dependence is negligible when $s$ is small compared to $1$; see \citealt{frank_2011}). We will refer to both the continuous and discrete time selection equations \eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} throughout this paper, but the simpler continuous time case will be our point of comparison for the rest of this section.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} Phase diagram for the densities of two types $n_1$ and $n_2$ undergoing selection. (a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_2>K_1$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

@@

In the special case that all types are competitively equivalent (identical $c_i$), Eq.~\eqref{eq:master} takes a simpler form,
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U=\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:masterequalc}
\end{equation}
where we have used the fact that $L=\overline{b}N/T$ to make the dependence on $b$ and $N$ explicit ($\overline{b}$ is the population mean $b$). This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.

Similarly, the total number of territories acquired is
\begin{equation}
\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N) \label{eq:Nmaster}
\end{equation}

\subsection*{Density regulation and selection in the variable-density lottery}

Equipped with Eq.~\eqref{eq:master} we now outline the basic properties of the $b$, $c$ and $d$ traits. Adult density $N$ is regulated by the birth and mortality rates $b$ and $d$; $b$ controls the fraction of unoccupied territories that are contested (see Eq.~\eqref{eq:Nmaster}), while $d$ controls adult mortality. Competitive ability $c$ does not enter Eq.~\eqref{eq:Nmaster}, and therefore does not regulate total adult density: $c$ only affects the relative likelihood of winning a contested territory.

Selection in our variable-density lottery model is in general density-dependent, by which we mean that the discrete-time selection factor $(W_2-W_1)/\overline{W}$ from Eq.~\eqref{eq:canonicaldisc} may depend on $N$. More specifically, as we show below, $b$ and $c$ selection are density dependent, but $d$ selection is not. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

The strength of $b$ selection declines with increasing density. When types differ in $b$ only ($b$-selection), Eq.~\eqref{eq:master} simplifies to Eq.~\eqref{eq:masterequalc}, and absolute fitness can be written as $W_i=(1+\frac{b_i}{\overline{b}}f(\overline{b},N))/d_i$ where $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is a decreasing function of $N$. Thus, the selection factor $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ declines with increasing density: the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

In the case of $c$-selection, Eq.~\eqref{eq:master} implies that $W_2-W_1$ is proportional to $\frac{T-N}{T} \left[(R_2+A_2)c_2-(R_1+A_1)c_1\right]/\overline{c}$. The strength of $c$-selection thus peaks at an intermediate density (Fig.~\ref{fig:DDS_lottery}), because most territories are claimed without contest at low density ($R_1,R_2,A_1,A_2\rightarrow 0$), whereas at high density few unoccupied territories are available to be contested ($T-N\rightarrow 0$).

Selection on $d$ is independent of density, because the density-dependent factor $1 + \frac{\Delta_+ n_i}{n_i}$ in Eq.~\eqref{eq:absfit} is the same for types that differ in $d$ only.

@@

We now turn to the issue of how density responds to selection in our variable-density lottery and previous models of selection in density-regulated populations. In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction). In our variable-density lottery model, however, the competitive ability trait $c$ is not density-regulating, even though $c$ contributes to fitness under crowded conditions. Consequently, $c$-selection does not cause density to change. In this section we compare this $c$-selection behavior with the previous literature, which we take to be exemplified by MacArthur's $K$-selection argument \citep{macarthur_1967}.

@@

In contrast to both $K$ and $\alpha$ selection, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once $N$ reaches demographic equilibrium, selective sweeps behave indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). Thus, for $c$ sweeps, the selection factor $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} depends on frequency only, not density, provided that nothing else pushes $N$ out of demographic equilibrium over the sweep. This uncoupling of density from ongoing $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.

@@

For the relative fitness model Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_2-W_1)/\overline{W}$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, $(W_2-W_1)/\overline{W}$ is independent of $N$ in the case of $d$-selection. Selection is also independent of $N$ when the population is at demographic equilibrium and $N$ is unaffected by ongoing selection; as is the case for $c$-selection. Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing. This can obviously occur if density-dependent selection happens in a population far from demographic equilibrium, in which case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonicaldisc} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait threatens Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models. This applies, for example, to $K$-selection in the logistic \citep{kimura1969natural,crow_1970}. We consider the simple birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i, \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}).

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{bsweep.pdf}
\caption{\label{fig:bsweep} Equilibrium $b$ sweeps behave as though selection is independent of density even though $b$-selection is density-dependent in general. Panel (b) shows the density-dependent selection factor $(W_2-W_1)/\overline{W}$ predicted by Eq.~\eqref{eq:master} (solid line) compared to the same selection factor with the density-dependence term $f(\overline{b},N)$ held constant at its initial value (dashed line).}
\end{figure}

However, if selection acts simultaneously on more than one trait in our variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on another trait subject to density-dependent selection. For instance, if selection acts simultaneously on $b$ and $d$, then $f(\overline{b},N)$ changes value from $d_1-1$ to $d_2-1$ over a sweep. The dynamics of density will then affect the selection factor $(W_2-W_1)/\overline{W}$ and cause deviations analogous to selection on $\delta$ in the continuous time case (Fig.~\ref{fig:bdsweep}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{bdsweep.pdf}
\caption{\label{fig:bdsweep} Simultaneous selection on $b$ and $d$ induces density-dependence in the selection factor $(W_2-W_1)/\overline{W}$. Panel (b) shows the predictions of Eq.~\eqref{eq:masterequalc} (solid line) versus the same with the density-dependence factor$f(\overline{b},N)$ held constant at its initial value.}
\end{figure}

@@

Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is effectively independent of density). This density-independence breaks down when strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:bdsweep}). The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}. The combination of both is necessary to threaten the constant-$s$ approximation. Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

Selection in the variable-density lottery is quite different from classical density-dependent selection (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection''). In the latter, only one life-history stage is represented, and the effects of crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation). Selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients ($\alpha$-selection). These are both ``$\delta$-like''  in the sense that selection depends on density and also causes density to change ($\delta$ is defined in Eq.~\eqref{eq:simplebirthdeath}). Strong selection is therefore sufficient for Eq.~\eqref{eq:canonicalcont} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.

@@

Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature. Even if reproductive excesses are ubquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent. For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent. However, this conclusion is sensititive to the assumptions of well-mixed resource competition models. Spatial localization of consumable resources (e.g. for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \citep{bolker_99}. In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).

Moreover, even in well mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference. Interference competition can dramatically alter the dynamics of resource exploitation \citep{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection. For instance, sexual selection can be viewed as a form of relative interference competition between genotypes. Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like. Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important. Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \citep{gill_1974}. However, $\alpha$-selection does transiently affect population density and therefore retains $\delta$-like features.

The above findings underscore that the most serious threat to the density-independent models of selection (Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicalcont}) arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e. as a result of a temporally-variable environment. While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the density-independent approximation, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold). By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig. 9.5 in \cite{otto_2011}; Fig. 5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there may indeed be no choice but to abandon relative fitness. Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2) lottery models in general are mathematically similar to the Wright-Fisher model, which should facilitate the analysis of genetic drift when $N$ is unstable.

@@

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types. To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, with low $l_1\ll 1$ and high density $l_2\approx L\gg 1$ respectively. Since $L$ is large, uncontested territories make up a negligible fraction of the total. The rare type's territorial acquisition is almost entirely due to $\Delta_r n_1$, while the common type's territorial acquisition entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This behavior is clearly pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.

@@

Eq.~\eqref{eq:cvr} reveals two key points. First, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation. In this case, the right hand side of Eq.~\eqref{eq:cvr} is approximately equal to $C^{1/2}\left(\frac{1}{L-l_i}+1-\frac{1}{C}\right)^{1/2}$, which is small for both low and high nonfocal densities. The worst case scenario occurs when $L-l_i$ is of order one, and it can be directly verified that Eq.~\eqref{eq:meanfieldr} is then still a good approximation (see Fig.~\ref{fig:approx_details}). Second, if some of the $c_j$ are much larger than the others, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large. Specifically, in the presence of a rare,  strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the right hand side of Eq. \eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}. Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when there are large differences in $c$.
